Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2017-06-26 paper/submitted/round2_png/mathematical_formulation_v4_cmame_vikram_HLedit.tex
notation edits

Extending the property in \Cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Following~\cite{BecVex05}, we rewrite the term $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$ using the Fundamental Theorem of Calculus,
%
\begin{equation}
\label{eq:Mantiderivative}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \int\limits_{0}^{1} \mathcal{M}'_{HF}\left(\Psi_{HF} + se\right)\left(e\right) \, \textrm{d}s.
\end{equation}
%
Applying the trapezoidal rule to the integral in~\Cref{eq:Mantiderivative} gives,
%
\begin{equation}
\label{eq:Mprimetrapezoid}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{HF})(e) + \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(e) + \mathcal{R}\textrm{.}
\end{equation}
%
where the remainder $\mathcal{R}$ is given as,
%
\begin{equation}
\label{eq:trapezoidremainder}
\mathcal{R} = \frac{1}{2} \int\limits_{0}^{1} \mathcal{M}'''_{HF}\left(\Psi_{LF} + se\right)(e,e,e) \, s(s-1) \, \textrm{d}s,
\end{equation}
%
which is finite due to the third order differentiability assumptions. We further note that $\mathcal{M}'_{HF}(\Psi_{HF})(e)$ vanishes since $\Psi_{HF}$ is a stationary point of $\mathcal{M}_{HF}$, and~\Cref{eq:Mprimetrapezoid} can be simplified to,
%
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}\textrm{.}
\end{equation}
%
Combining \Cref{eq:repIwithM} and \Cref{eq:beckvex} we obtain,
%
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}\textrm{,}
\end{multline}
which completes the proof.
\end{proof}
%

@@

%
\begin{proposition}
\label{thm:error_estimate_dual}
Let the hypotheses of Proposition~\ref{thm:error_estimate} hold. Let $\Lambda$ be the solution of the dual problem, \red{$\mathcal{M}'$ has just one '?}
%
\begin{equation}
\mathcal{M}_{HF}'(\Psi_{LF};\Phi, \Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2,
\end{equation}
%
where $\mathcal{Q}:(U_{HF} \times U_{HF} \times Q_{HF})^2 \to \R$ is a linear functional defined by,
%
\begin{equation}
\mathcal{Q}(\Phi)=\mathcal{M}'_{HF}(\Psi_{LF})(\Phi).
\label{eq:supadjout}
\end{equation}
%
Then we have,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF}) + \mathcal{R}_{\mathscr{R}} + \mathcal{R}
\end{multline}
%
where $\mathcal{R}_{\mathscr{R}}$ is a second-order term in the error $e=\Psi_{HF}-\Psi_{LF}$, and $\mathcal{R}$ is as defined in \Cref{thm:error_estimate}. \red{Now that we've explicitly pointed out the second-order error term, can we really say that we derived "an adjoint-based third-order estimate" (in abstract and intro)?}
%
\end{proposition}
%
\begin{proof}
%
The proof simply follows from the classic dual-based a posteriori error estimation strategy. Note that, as a stationary point, $\Psi_{HF}$ satisfies,
%
\begin{equation}
\mathscr{R}(\Psi_{HF},\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys}
\end{equation}
%
which is represents the equation $\mathcal{M}'_{HF}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define the corresponding adjoint problem
%
\begin{equation}
\mathscr{R}'(\Psi_{LF};\Phi,\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
%
for the supplementary adjoint $\Lambda$. The error in the linear output $\mathcal{Q}$ defined in \Cref{eq:supadjout} can then be expressed as a dual-weighted residual,
%
\begin{equation}
\label{eq:adjOutErr}
\mathcal{Q}(\Psi_{HF}-\Psi_{LF})=-\mathscr{R}(\Psi_{LF},\Lambda) + R_{\mathscr{R}},
\end{equation}
%
where $R_{\mathscr{R}}$ is a second order remainder term due to the linearization of $\mathscr{R}$ about $\Psi_{LF}$. This implies the required result. \red{What is the classical notation for adjoint problems? Is it okay that we're breaking from the previous notation where a second pair of parentheses denotes things that we are linear in?}
%
\end{proof}
%
Other approaches to estimating the term $\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})$ can become feasible in certain scenarios; for example, if the high- and low-fidelity models only differ in the computational grid used, the difference $\Psi_{HF}-\Psi_{LF}$ can be approximated using interpolation or patch recovery methods~\cite{BecVex05}.

>>>>

Extending the property in \Cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Following~\cite{BecVex05}, we rewrite the term $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$ using the Fundamental Theorem of Calculus,
%
\begin{equation}
\label{eq:Mantiderivative}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \int\limits_{0}^{1} \mathcal{M}'_{HF}\left(\Psi_{HF} + se\right)\left(e\right) \, \textrm{d}s.
\end{equation}
%
Applying the trapezoidal rule to the integral in~\Cref{eq:Mantiderivative} gives,
%
\begin{equation}
\label{eq:Mprimetrapezoid}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{HF})(e) + \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(e) + \mathcal{R}\textrm{.}
\end{equation}
%
where the remainder $\mathcal{R}$ is given as,
%
\begin{equation}
\label{eq:trapezoidremainder}
\mathcal{R} = \frac{1}{2} \int\limits_{0}^{1} \mathcal{M}'''_{HF}\left(\Psi_{LF} + se\right)(e,e,e) \, s(s-1) \, \textrm{d}s,
\end{equation}
%
which is finite due to the third-order differentiability assumptions. We further note that $\mathcal{M}'_{HF}(\Psi_{HF})(e)$ vanishes since $\Psi_{HF}$ is a stationary point of $\mathcal{M}_{HF}$, and~\Cref{eq:Mprimetrapezoid} can be simplified to,
%
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}\textrm{.}
\end{equation}
%
Combining \Cref{eq:repIwithM} and \Cref{eq:beckvex} we obtain,
%
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}\textrm{,}
\end{multline}
which completes the proof.
\end{proof}
%

@@

%
\begin{proposition}
\label{thm:error_estimate_dual}
Let the hypotheses of Proposition~\ref{thm:error_estimate} hold. Let the variational form
%
\begin{equation}
\mathscr{R}(\Psi_{HF})(\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys_prop}
\end{equation}
%
represent the equation $\mathcal{M}'_{HF,\Psi}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define an adjoint problem
%
\begin{equation}
\mathscr{R}'(\Psi_{LF},\Phi)(\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
for the supplementary adjoint $\Lambda$, where $\mathscr{R}$ is defined in \Cref{eq:supadjsys_prop}, and where
%Let $\Lambda$ be the solution of the dual problem, \red{$\mathcal{M}'$ has just one '?}
%
%\begin{equation}
%\mathcal{M}_{HF}'(\Psi_{LF};\Phi, \Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2,
%\end{equation}
%
where $\mathcal{Q}:(U_{HF} \times U_{HF} \times Q_{HF})^2 \to \R$ is a linear functional defined by,
%
\begin{equation}
\mathcal{Q}(\Phi)=\mathcal{M}'_{HF}(\Psi_{LF})(\Phi).
\label{eq:supadjout}
\end{equation}
%
Then we have,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF}) + \mathcal{R}_{\mathscr{R}} + \mathcal{R}
\end{multline}
%
where $\mathcal{R}_{\mathscr{R}}$ is a second-order term in the error $e=\Psi_{HF}-\Psi_{LF}$, and $\mathcal{R}$ is as defined in \Cref{thm:error_estimate}. %\red{Now that we've explicitly pointed out the second-order error term, can we really say that we derived "an adjoint-based third-order estimate" (in abstract and intro)?}
%
\end{proposition}
%
\begin{proof}
%
The proof simply follows from the classic dual-based a posteriori error estimation strategy. Note that, as a stationary point, $\Psi_{HF}$ satisfies,
%
\begin{equation}
\mathscr{R}(\Psi_{HF})(\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys}
\end{equation}
%
which is represents the equation $\mathcal{M}'_{HF}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define the adjoint problem
%
\begin{equation}
\mathscr{R}'(\Psi_{LF},\Phi)(\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
%
for the supplementary adjoint $\Lambda$. The error in the linear output $\mathcal{Q}$ defined in \Cref{eq:supadjout} can then be expressed as a dual-weighted residual,
%
\begin{equation}
\label{eq:adjOutErr}
\mathcal{Q}(\Psi_{HF}-\Psi_{LF})=-\mathscr{R}(\Psi_{LF})(\Lambda) + R_{\mathscr{R}},
\end{equation}
%
where $R_{\mathscr{R}}$ is a second order remainder term due to the linearization of $\mathscr{R}$ about $\Psi_{LF}$ in \Cref{eq:superAdjEq}. This implies the required result. %\red{What is the classical notation for adjoint problems? Is it okay that we're breaking from the previous notation where a second pair of parentheses denotes things that we are linear in?}
%
\end{proof}
%
Other approaches to estimating the term $\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})$ can become feasible in certain scenarios; for example, if the high- and low-fidelity models only differ in the computational grid used, the difference $\Psi_{HF}-\Psi_{LF}$ can be approximated using interpolation or patch recovery methods~\cite{BecVex05}.

### 2017-06-26 paper/submitted/round2_png/numerical_experiments_v4_cmame.tex
notation edits

%------------------------------------------------------------------------------------------------------------------------%
\subsection{Variable Physics: Convection-Diffusion(-Reaction)} \label{sec:cdvcdr}
%------------------------------------------------------------------------------------------------------------------------%
In this section, we consider a pair of models which differ in the physics included. In \Cref{sec:cdvcdrSetup} we describe a baseline setup for a simple two-dimensional problem. \Cref{sec:cdvcdrBaseRef} describes the results of applying \Cref{alg:refSeries} to the baseline problem, and \Cref{sec:qoivdata} describes the results of changing the placement of the observations or the QoI region from the baseline.
%
%------------------------------------------------------------%
\subsubsection{Problem Setup} \label{sec:cdvcdrSetup}
%------------------------------------------------------------%
%
We consider a rectangular domain $\Omega(x_1,x_2)=[0,5]\times[0,1]$, where $x_1$ and $x_2$ are the spatial coordinates. The high-fidelity model is a single-species convection-diffusion-reaction equation with a nonlinear reaction term, described by,
%
\begin{subequations}
\label{eq:cdvcdrHF}
\begin{align}
k_d\nabla^2 u - \vec{V}\cdot\nabla u + k_ru^2 = f(q) \quad &\text{in } \Omega, \label{eq:cdvcdrHF_int} \\
u = 0 \quad &\text{on } \partial \Omega, \label{eq:cdvcdrHF_bdry}
\end{align}
\end{subequations}
%
where the state $u$ is the species concentration and $f(q)$ is a forcing field described by the parameters. We have a divergence-free parabolic-profile velocity field $\vec{V}(x_1,x_2) = (2x_2(1-x_2),0)$; the diffusion and reaction coefficients are $k_d = 0.1$ and $k_r = -42.0$, respectively. The low-fidelity model,
%
\begin{equation}
k_d\nabla^2 u - \vec{V}\cdot\nabla u = f(q)
\end{equation}
%
differs only in the removal of the reaction term. To form the mixed-fidelity models, we divide the domain into complementary subdomains, $\Omega_{HF}$ and $\Omega_{LF}$, where the high- and low-fidelity models are solved, respectively. The resulting mixed-fidelity models can be described by,
%
\begin{equation}
k_d\nabla^2 u - \vec{V}\cdot\nabla u + k^{MF}_ru^2= f(q),
\end{equation}
%
where $k^{MF}_r$ is a piecewise-constant reaction coefficient,
%
\begin{equation}
k^{MF}_r=
\begin{cases}
k_r & \textrm{if }x\in\Omega_{HF} \\
0 & \textrm{if }x\in\Omega_{LF}.
\end{cases}
\end{equation}
%
Homogeneous Dirichlet boundary conditions are applied on the entire boundary of the domain. The QoI we wish to calculate is the integral of the state,
%
\begin{equation}
I(q,u)=\int_{(x_1,x_2)\in \Omega_I} u \:\textrm{d}A,
\end{equation}
%
over a region $\Omega_I=[0.625,0.875]\times[0.375,0.625]$.

The unknown parameters to be inferred correspond to the forcing field, so that in this case $f(q)=q$. The observations are sparse measurements of the state.
\Cref{fig:baseSetup} shows the locations of the observations and the region $\Omega_I$ over which the QoI is calculated. For the low-fidelity model, the inverse problem is linear (i.e., the observations depend linearly on the parameters). Since the inverse problem is ill-posed, we use Tikhonov regularization~\cite{EngHanNeu00}. The regularization term in the objective function \Cref{eq:invOpt_obj} is $R(q)=\frac{\beta}{2}\int_\Omega \|\nabla f(q)\|_2^2\:\textrm{d}A$, where $\beta=10^{-5}$ is the regularization coefficient; in this work, $\beta$ is chosen arbitrarily, though methods exist for choosing optimal regularization coefficients~\cite{BelgeKilmerMiller02,Calvetal00}.
%
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/setup_3_3.pdf}
\caption{Locations of the observations and the QoI region.}
\label{fig:baseSetup}
\end{figure}
%

>>>>

%------------------------------------------------------------------------------------------------------------------------%
\subsection{Variable Physics: Convection-Diffusion(-Reaction)} \label{sec:cdvcdr}
%------------------------------------------------------------------------------------------------------------------------%
In this section, we consider a pair of models which differ in the physics included. In \Cref{sec:cdvcdrSetup} we describe a baseline setup for a simple two-dimensional problem. \Cref{sec:cdvcdrBaseRef} describes the results of applying \Cref{alg:refSeries} to the baseline problem, and \Cref{sec:qoivdata} describes the results of changing the placement of the observations or the QoI region from the baseline.
%
%------------------------------------------------------------%
\subsubsection{Problem Setup} \label{sec:cdvcdrSetup}
%------------------------------------------------------------%
%
We consider a rectangular domain $\Omega(x_1,x_2)=[0,5]\times[0,1]$, where $x_1$ and $x_2$ are the spatial coordinates. Recall the pair of models introduced in \Cref{sec:adapt_alg}. The high-fidelity model is a single-species convection-diffusion-reaction equation with a nonlinear reaction term, described by \Cref{eq:cdvcdrHF}. The low-fidelity model, described by \Cref{eq:cdvcdrLF}, differs only in the removal of the reaction term. We have a divergence-free parabolic-profile velocity field $\vec{V}(x_1,x_2) = (2x_2(1-x_2),0)$; the diffusion and reaction coefficients are $k_d = 0.1$ and $k_r = -42.0$, respectively. Homogeneous Dirichlet boundary conditions are applied on the entire boundary of the domain.

The QoI region is $\Omega_I=[0.625,0.875]\times[0.375,0.625]$. The parameters $q$ we wish to infer correspond to the forcing function. The observations are sparse measurements of the state.
\Cref{fig:baseSetup} shows the locations of the observations and the region $\Omega_I$ over which the QoI is calculated. Note that for the low-fidelity model, the inverse problem is linear (i.e., the observations depend linearly on the parameters). We arbitarily choose the regularization coefficient to be $\beta=10^{-5}$; in this work, $\beta$ is chosen arbitrarily, though methods exist for choosing optimal regularization coefficients~\cite{BelgeKilmerMiller02,Calvetal00}.
%
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{baseSeries/setup_3_3.pdf}
\caption{Locations of the observations and the QoI region.}
\label{fig:baseSetup}
\end{figure}
%
